Nail: A practical interface generator for data formats

### 2014-04-01 paper/design.tex
rough intro pass done, with some sizable holes left
Intro

\section{Design}

\paragraph{Overview}
To alleviate these problems, Nail provides a richer set of combinators that induce the structure of
an internal model, while also describing the grammar of the external protocol. Nail takes the
combinators and produces type declarations for an internal model, the \textit{parser}, a function to parse a sequence of
bytes into an instance of the model and the \textit{generator}, a function to create a sequence of
bytes from an instance of the model.

A central design decision of Nail is that there is a  bijection between the model exposed to the programmer and the bytes
input and output up to syntactic equivalence. More precisely, the parser is the generators inverse,
so parsing the generators output will yield the generators input, but generating the parsers output
does not necessarily yield the parsers input. To understand why this makes sense, consider a grammar
for a text language that tolerates white space or a binary protocol that tolerates arbitrarily long
padding\footnote{Say, the physical layer of most communication protocols is a possibly infinite
  sequence of symbols that are syntactically nil followed by a pre-determined synchronisation
  sequence and the actual contents of the transmission.}. However, if some features of Nail are
avoided, there is a proper isomorphism between model and protocol.

In the following, we present the combinators in Nail grouped by the combinators of conventional Parser Expression Grammars they replace.
\paragraph{Fundamental parsers}
The elementary parsers of Nail are the same as those of Hammer, signed and unsigned integers with
arbitrary lengths up to 64 bits. Note that is possible to define parsers for sub-byte lengths, e.g.
to parse the 4-bit data offset within the TCP header. Integer parsers return their value in the
smallest appropriately sized machine integer type, e.g. a 24 bit integer is stored in a 32-bit wide variable.\\
Examples: \texttt{uint4}, \texttt{int32}

>>>>

\section{Design}
\label{s:design}

\paragraph{Overview}

To alleviate these problems, Nail provides a richer set of combinators that not only describes the
grammar of the external protocol, but also induce an internal model.

A central design decision of Nail is that there is a semantic bijection between the model exposed to
the programmer and the bytes input and output up to syntactic equivalence. More precisely, the
parser is the generators inverse, so parsing the generators output will yield the generators input,
but generating the parsers output does not necessarily yield the parsers input. To understand why
this makes sense, consider a grammar for a text language that tolerates white space or a binary
protocol that tolerates arbitrarily long padding\footnote{Say, the physical layer of most
  communication protocols is a possibly infinite sequence of symbols that are syntactically nil
  followed by a pre-determined synchronisation sequence and the actual contents of the
  transmission.}. In that case, the program semantics should be independent of the number of padding
elements in the input, and Nail therefore does not expose that information to the programmer. We
call such discarded fields \emph{constants}. Currently, Nail always makes a default choice when
there are multiple options to express a constant, however Nail could be extended to allow a
grammar-specific plug-in to make these choices, say for faster alignment, streaming applications
when data is not ready or visual appearance in human-readable protocols. However, if some features
of Nail are avoided, there is a proper isomorphism between model and protocol.

Nail evaluates the combinators and produces type declarations for the internal model, the \textit{parser}, a function to parse a sequence of
bytes into an instance of the model and the \textit{generator}, a function to create a sequence of
bytes from an instance of the model.

In the following, we present the combinators in Nail.
\paragraph{Fundamental parsers}
The elementary parsers of Nail are the same as those of Hammer, signed and unsigned integers with
arbitrary lengths up to 64 bits. Note that is possible to define parsers for sub-byte lengths, e.g.
to parse the 4-bit data offset within the TCP header. Integer parsers return their value in the
smallest appropriately sized machine integer type, e.g. a 24 bit integer is stored in a 32-bit wide variable.\\
Examples: \texttt{uint4}, \texttt{int32}

### 2014-04-01 paper/intro.tex
Intro

This paper presents the initial design and implementation of Nail, a parser
generator that greatly reduces the programmer effort required to use a
grammar-based parser. Nail addresses the above three challenges with several key
ideas. First, Nail reduces the expressiveness of its grammar language by
removing semantic actions. Existing parser generators allow arbitrary
computation to transform between the AST and the parser output.%  Instead, Nail
% derives the structure of its output automatically from the grammar, forcing the
% programmer to clearly separate syntactic validation and semantic processing.

Second, this allows the parser to construct a \emph{syntactic bijection} between input
and output. That is, parsing is the inverse of generating output. However,
Nail allows \emph{constants} in the external format to have multiple
representations if they should not affect the semantics of the data. For
example, in a text protocol, the amount of white-space separating tokens should
not affect the meaning of the data and is consequently not exposed to the
application by Nail in the internal representation. However, as long as
constants are only used for their intended purpose of representing syntax-only
features, the generated output will have the same semantics as the parsed input.

Third, Nail provides support for structural dependencies, transparently handling
offset and length fields. XXX todo

>>>>

This paper presents the initial design and implementation of Nail, a parser
generator that greatly reduces the programmer effort required to use a
grammar-based parser. Nail addresses the above three challenges with several key
ideas.

First, Nail reduces the expressiveness of its grammar language by removing
semantic actions. Existing parser generators allow arbitrary computation to
transform between the AST and the parser output. Instead, Nail derives the
structure of its output automatically from the grammar, forcing the programmer
to clearly separate syntactic validation and semantic processing.

Second, this well-defined internal representation allows Nail to inverse the
parser and generate output. However, Nail allows \emph{constants} in the
external format to have multiple representations if they should not affect the
semantics of the data. For example, in a text protocol, the amount of
white-space separating tokens should not affect the meaning of the data and
consequently Nail does not expose it to the application. As long as
constants are only used for their intended purpose of representing syntax-only
features, the generated output will have the same semantics as the parsed input.

Third, Nail provides support for structural dependencies, transparently handling
offset and length fields. By hiding the offset and length values from the
programmer, Nail ensures they remain consistent even if the data is changed by
the application.
